---
title: "FFPE_imputation"
author: "Fares Burwag"
date: "30/06/2023"
output: html_document
---

```{r setup}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
# library('iq')
library('RColorBrewer')
library('pheatmap')
library("limma")
library("pcaMethods",mask.ok = F)
library("missForest")
library("imputeLCMD")
library("foreach")
library("doParallel")
library("grid")
library("gridExtra")

```

Only run the chunk below if you do not have the reportParsedIqQuant.tsv file generated. It takes a while to run...

```{r}
##########################################################################################
##first we can filter and parse the report data into a better format for IQ
raw = read_tsv('../fullFFPECohortHSJBCC/report.tsv', show_col_types = FALSE) %>%
  dplyr::mutate(File.Name = sub('.*faresFfpeTissue[A-Za-z]*_(.*)_dia_[123]\\.raw', '\\1', File.Name)) %>%
  dplyr::filter(Q.Value <= 0.01)

##save the data
write.table(raw, '../reportParsed.tsv', 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')


##run the IQ analysis
iqResult = process_long_format('../reportParsed.tsv', 
                           output_filename = '../reportParsedIqQuant.tsv',
                           annotation_col = 'Genes',
                           normalization = 'median',
                           filter_double_less = c("Q.Value" = "0.01", "PG.Q.Value" = "0.01"),
                           peptide_extractor = function(x) gsub("[0-9].*$", "", x))

##parse the IQ result
proData = read_tsv('../reportParsedIqQuant.tsv', show_col_types = FALSE) %>%
  dplyr::rename(accession = 'Protein.Group',
                symbol = 'Genes')


##save the data
saveRDS(proData,file = "./parsedRDSFile.rds")
write.table(proData, "processedRDS.rds.tsv",
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
```

Start running code here unless you want to regenerate protein report file.

```{r}
# Reading in parsed report file
proData <- read_rds("parsedRDSFile.rds")


proData_withMissing <- proData[,5:ncol(proData)] %>%
  as.data.frame()

## Remove samples with excessive missing values (only keep samples where > 50% of genes are detected)
naFreqperSample <- apply(is.na(proData_withMissing), 2, function(x){sum(!x)/length(x)})
naFreqperSample_names <- names(naFreqperSample)
naFreqperSample_values <- unname(naFreqperSample)
naFreqperSample_threshold <- naFreqperSample_names[naFreqperSample_values >= 0.65]

row.names(proData_withMissing) <- proData$accession
proData_withMissing <- proData_withMissing[,naFreqperSample_threshold]

## Remove any genes where all samples have NA. Necessary for BPCA to work
proData_withMissing <- proData_withMissing[rowSums(!is.na(proData_withMissing)) > 0,]

```

```{r}


knnTn_proDatawM <- proData_withMissing  %>%
  rownames_to_column("accession")

# To create a validation dataset to determine best K, lets introduce a few
# artifical NAs in the lowest 10% proteins in each sample
knnTn_proDataSimulatedNAs <- knnTn_proDatawM
denominator <- sum(knnTn_proDataSimulatedNAs[,-1] %>% apply(1, function(x) sum(!is.na(x)) >= 5))
knnTn_proDataSimulatedNAs[,-1] <- knnTn_proDataSimulatedNAs[,-1] %>% apply(2, function(x) replace(x, sample(which((!is.na(x) & x<=quantile(x, 0.05, na.rm = TRUE, names = FALSE))),size = 100), NA))
numerator <- sum(knnTn_proDataSimulatedNAs[,-1] %>% apply(1, function(x) sum(!is.na(x)) >= 5))
numerator/denominator

knnTn_proDataSimulatedNAs <- knnTn_proDataSimulatedNAs[knnTn_proDataSimulatedNAs[,-1] %>% apply(1, function(x) sum(!is.na(x)) >= 5),]


```

Naive methods that infer no information from global or local data structures (simply use a single constant or a randomly generated value)

```{r}

# Testing out naive methods for imputation - NA to 0, SampMin, and MinProb
naive_proDatawM <- knnTn_proDataSimulatedNAs

# NA's to 0
toZero_naiveSupplemented_proData <- naive_proDatawM
toZero_naiveSupplemented_proData[is.na(toZero_naiveSupplemented_proData)] <- 0

# NA's to SampMin
sampMin_naiveSupplemented_proData <- naive_proDatawM
sampMin_naiveSupplemented_proData[,-1] <- apply(sampMin_naiveSupplemented_proData[,-1], 2,
                                                function(x) ifelse(is.na(x), min(x, na.rm = T), x))

# NA's to minProb (select randomly form lowest 1% of data)
minProb_naiveSupplemented_proData <- naive_proDatawM
minProb_naiveSupplemented_proData[,-1] <- apply(minProb_naiveSupplemented_proData[,-1], 2,
                                                function(x) replace(x, is.na(x),sample(x[!is.na(x) & x<=quantile(x, 0.01, na.rm = TRUE, names = FALSE)],sum(is.na(x)), replace = TRUE)))

```

```{r}

### Testing out KNN-TN to impute

# Assuming we want to test out k = 1 to k = 7, we need to select genes with at 
# least 5 nonNA samples. To be safe, lets set threshold at 10

tempTable <- knnTn_proDatawM[knnTn_proDatawM$accession %in% knnTn_proDataSimulatedNAs$accession,]
length(tempTable[is.na(knnTn_proDataSimulatedNAs) & !is.na(tempTable)])
# There's around 7000 datapoints to test from... good enough.

# shuffledKnnTn_proDataSimulatedNAs <- knnTn_proDataSimulatedNAs[sample(1:nrow(knnTn_proDataSimulatedNAs), nrow(knnTn_proDataSimulatedNAs), replace = FALSE),]
# shuffledKnnTn_proDataSimulatedNAs$set <- c(rep(seq(1, 4, 1), floor(nrow(shuffledKnnTn_proDataSimulatedNAs)/4)), rep(NA, 3))
# shuffledKnnTn_proDataSimulatedNAs <- shuffledKnnTn_proDataSimulatedNAs[!is.na(shuffledKnnTn_proDataSimulatedNAs$set),]

# This algorithm takes ages to run, therefore we'll select a k-value from 1 to 7 using cross validation with a split of 4.
knnTn_performance <- tibble(k = as.numeric(NA), rmspe = as.numeric(NA), dataTibble = as.data.frame(NA))
stored_dataPerformance <- list()
for (k_val in seq(2, 25, 2)){
  print(k_val)
  loop_data <- knnTn_proDataSimulatedNAs
  loop_data[,-1] <- imputeKNN(data = loop_data[,-1] %>% as.matrix(), k = k_val, distance = "truncation")
  
  completeData <- knnTn_proDatawM[knnTn_proDatawM$accession %in% loop_data$accession,]
  predicted_values <- loop_data[is.na(knnTn_proDataSimulatedNAs) & !is.na(completeData)]
  accurate_values <- completeData[is.na(knnTn_proDataSimulatedNAs) & !is.na(completeData)]

  # Get accuracy
  rmspe_value <- sqrt(mean((as.numeric(predicted_values)-as.numeric(accurate_values))^2))

  knnTn_performance <- bind_rows(knnTn_performance, tibble(k = k_val, rmspe = rmspe_value))
  stored_dataPerformance <- append(stored_dataPerformance, loop_data)
}

knnTn_supplementedData <- knnTn_proDataSimulatedNAs
knnTn_supplementedData[,-1] <- imputeKNN(data = knnTn_proDataSimulatedNAs[,-1] %>% as.matrix(), k = 15, distance = "truncation")

## Use k = 20 - generally performs better with higher number of neighbors

unique(knnTn_performance$rmspe)
sd(completeData$BCC16, na.rm = T)
hist((as.numer1ic(predicted_values)-as.numeric(accurate_values))^2,breaks = 1000)
boxplot(as.numeric(proData_withMissing$BCC1), as.numeric(accurate_values), as.numeric(predicted_values))

plot(x = knnTn_performance$k,y = knnTn_performance$rmspe)

```


```{r}

bpca_proDataSimulatedNAs <- knnTn_proDataSimulatedNAs
bpca_proDataSimulatedNAs <- log2(bpca_proDataSimulatedNAs[,-1])

# Using Bayesian PCA to infer value of missing data based on global patterns in dataset - Takes a few minutes to run
pc <- pca(bpca_proDataSimulatedNAs, nPcs = 10, method = "bpca")
proData_supplemented <- completeObs(pc)
proData_supplemented <- as.data.frame(proData_supplemented) %>%
  map_df(function(x) 2^x)


```

```{r}

# random forest
randomForest_proDataSimulatedNAs <- knnTn_proDataSimulatedNAs
randomForest_proDataSimulatedNAs[,-1] <- as.data.frame(log2(randomForest_proDataSimulatedNAs[,-1]))

registerDoParallel(cores=8)
rF_output <- missForest(randomForest_proDataSimulatedNAs[,-1], parallelize = "forests")

rF_output <- rF_output$ximp
rF_output_values <- as.data.frame(2^rF_output)


```
Visualizations to characterize the output of each imputation method

```{r}

## to Zero visualization

allData_supplementedNAs <- knnTn_proDataSimulatedNAs %>%
  pivot_longer(cols = 2:38, names_to = "sample", values_to = "intensity")
# toZero_naiveSupplemented_proData[,-1][!(is.na(knnTn_proDataSimulatedNAs[,-1]) & !is.na(completeData[,-1]))]

toZero_VisData <- toZero_naiveSupplemented_proData[,-1]
toZero_VisData[!(is.na(knnTn_proDataSimulatedNAs[,-1]))] <- NA
toZero_VisData <- toZero_VisData %>%
  pivot_longer(cols = 1:37, names_to = "sample", values_to = "intensity") %>%
  na.omit()

toZeroVis <- ggplot()+
  geom_boxplot(data = allData_supplementedNAs, aes(x = sample, y = intensity))+
  geom_point(data = toZero_VisData, aes(x = sample, y = intensity), color = "red", position = "jitter", alpha = 0.002)+
  theme(axis.text.x = element_text(angle = 45))+
  labs(x = "Sample", y = "Signal Intensity", title = "toZero Imputation")

## to sample minimum visualization

sampMin_VisData <- sampMin_naiveSupplemented_proData[,-1]
sampMin_VisData[!(is.na(knnTn_proDataSimulatedNAs[,-1]))] <- NA
sampMin_VisData <- sampMin_VisData %>%
  pivot_longer(cols = 1:37, names_to = "sample", values_to = "intensity") %>%
  na.omit()

toSampMinVis <- ggplot()+
  geom_boxplot(data = allData_supplementedNAs, aes(x = sample, y = intensity))+
  geom_point(data = sampMin_VisData, aes(x = sample, y = intensity), color = "red", position = "jitter", alpha = 0.002)+
  theme(axis.text.x = element_text(angle = 45))+
  labs(x = "Sample", y = "Signal Intensity", title = "sampleMinimum Imputation")

## to minProb visualization

toMinProb_VisData <- minProb_naiveSupplemented_proData[,-1]
toMinProb_VisData[!(is.na(knnTn_proDataSimulatedNAs[,-1]))] <- NA
toMinProb_VisData <- toMinProb_VisData %>%
  pivot_longer(cols = 1:37, names_to = "sample", values_to = "intensity") %>%
  na.omit()

toProbMinVis <- ggplot()+
  geom_boxplot(data = allData_supplementedNAs, aes(x = sample, y = intensity))+
  geom_point(data = toMinProb_VisData, aes(x = sample, y = intensity), color = "red", position = "jitter", alpha = 0.002)+
  theme(axis.text.x = element_text(angle = 45))+
  labs(x = "Sample", y = "Signal Intensity", title = "minProbability Imputation")


## knnTruncation visualization

knnTn_VisData <- knnTn_supplementedData[,-1]
knnTn_VisData[!(is.na(knnTn_proDataSimulatedNAs[,-1]))] <- NA
knnTn_VisData <- knnTn_VisData %>%
  pivot_longer(cols = 1:37, names_to = "sample", values_to = "intensity") %>%
  na.omit()

toKnnTnVis <- ggplot()+
  geom_boxplot(data = allData_supplementedNAs, aes(x = sample, y = intensity))+
  geom_point(data = knnTn_VisData, aes(x = sample, y = intensity), color = "red", position = "jitter", alpha = 0.005)+
  theme(axis.text.x = element_text(angle = 45))+
  labs(x = "Sample", y = "Signal Intensity", title = "kNN Truncation Imputation")

## BPCA visualization
bpca_Supplemented_proData <- bind_cols(accession = knnTn_proDataSimulatedNAs$accession, proData_supplemented)
bpca_VisData <- bpca_Supplemented_proData[,-1]
bpca_VisData[!(is.na(knnTn_proDataSimulatedNAs[,-1]))] <- NA
bpca_VisData <- bpca_VisData %>%
  pivot_longer(cols = 1:37, names_to = "sample", values_to = "intensity") %>%
  na.omit()

tobpcaVis <- ggplot()+
  geom_boxplot(data = allData_supplementedNAs, aes(x = sample, y = intensity))+
  geom_point(data = bpca_VisData, aes(x = sample, y = intensity), color = "red", position = "jitter", alpha = 0.005)+
  theme(axis.text.x = element_text(angle = 45))+
  labs(x = "Sample", y = "Signal Intensity", title = "BPCA Imputation")

## randomForest visualization

randomForest_Supplemented_proData <- bind_cols(accession = knnTn_proDataSimulatedNAs$accession, rF_output_values)
randomForest_VisData <- randomForest_Supplemented_proData[,-1]
randomForest_VisData[!(is.na(knnTn_proDataSimulatedNAs[,-1]))] <- NA
randomForest_VisData <- randomForest_VisData %>%
  pivot_longer(cols = 1:37, names_to = "sample", values_to = "intensity") %>%
  na.omit()

toRFVis <- ggplot()+
  geom_boxplot(data = allData_supplementedNAs, aes(x = sample, y = intensity))+
  geom_point(data = randomForest_VisData, aes(x = sample, y = intensity), color = "red", position = "jitter", alpha = 0.005)+
  theme(axis.text.x = element_text(angle = 45))+
  labs(x = "Sample", y = "Signal Intensity", title = "randomForest Imputation")

ggsave(filename = "imputationResultsVis.png",
       plot = gridExtra::grid.arrange(toZeroVis, toSampMinVis, toProbMinVis, toKnnTnVis, tobpcaVis, toRFVis, nrow = 6),
       width = 7,
       height = 35)

```


```{r}

real_values <- completeData[is.na(knnTn_proDataSimulatedNAs) & !is.na(completeData)]
toZero_predicted_values <- toZero_naiveSupplemented_proData[is.na(knnTn_proDataSimulatedNAs) & !is.na(completeData)]
toSampMin_predicted_values <- sampMin_naiveSupplemented_proData[is.na(knnTn_proDataSimulatedNAs) & !is.na(completeData)]
toMinProb_predicted_values <- minProb_naiveSupplemented_proData[is.na(knnTn_proDataSimulatedNAs) & !is.na(completeData)]
tokNNTn_predicted_values <- knnTn_supplementedData[is.na(knnTn_proDataSimulatedNAs) & !is.na(completeData)]
toBPCA_predicted_values <- bpca_Supplemented_proData[is.na(knnTn_proDataSimulatedNAs) & !is.na(completeData)]
toRF_predicted_values <- randomForest_Supplemented_proData[is.na(knnTn_proDataSimulatedNAs) & !is.na(completeData)]

imputePerformance <- tibble(method_impute = c("Zero", "SampMin", "MinProb", "kNN Truncation", "BPCA", "Random Forest"),
       RMSPE_value = c(sqrt(mean(c(as.numeric(toZero_predicted_values) - as.numeric(real_values))^2)),
                       sqrt(mean(c(as.numeric(toSampMin_predicted_values) - as.numeric(real_values))^2)),
                       sqrt(mean(c(as.numeric(toMinProb_predicted_values) - as.numeric(real_values))^2)),
                       sqrt(mean(c(as.numeric(tokNNTn_predicted_values) - as.numeric(real_values))^2)),
                       sqrt(mean(c(as.numeric(toBPCA_predicted_values) - as.numeric(real_values))^2)),
                       sqrt(mean(c(as.numeric(toRF_predicted_values) - as.numeric(real_values))^2))
                       ))

ggplot(data = imputePerformance)+
  geom_bar(aes(x = reorder(method_impute, RMSPE_value), y = RMSPE_value), stat = "identity", fill = "salmon")+
  labs(x = "Imputation Method", y = "Root Mean Square Error")

ggsave("imputationPerformance.png")

```


For BC Childrens, the label key is as below

```{r}

BCC_key <- data.frame(short_label = c("BCC1", "BCC2", "BCC3", "BCC4", "BCC5", "BCC6", "BCC7", "BCC8", "BCC9", "BCC10", "BCC11", "BCC12", "BCC13", "BCC14", "BCC15", "BCC16", "BCC17", "BCC18", "BCC19", "BCC20", "BCC21", "BCC22", "BCC23", "BCC24", "BCC25", "BCC26", "BCC27", "BCC28", "BCC29", "BCC30", "BCC31", "BCC32", "BCC33", "BCC34", "BCC35", "BCC36"), original = c("S211165D1","S21335A2", "S20749B1", "S203243C1","S212984B1", "S212818B1","S12469D1","S123607B1","S203113B2","S194826A2","S202667C10","S20228B1","S172944B3","S193365A2","S181477A2","S184058A2","S171999C1","S161352A2","S16686C1","S16486C4","S16316A2","S154448C1","S152968D2","S184465A1","S151916B2","S144474A3","S22618A2","S144473A2","S134449A2","S134324C1","S131864D1","S125789A1","S123944D5","S122751B1","S122371D2","S121B5")) %>%
  mutate(original_dashed = sub("(S[0-9]{2})(.*)","\\1-\\2", original))

batch_key <- data.frame(sample = c(unique(colnames(proData)[5:ncol(proData)]))) %>%
  mutate(batch = case_when(grepl("hsj", sample, ignore.case = F) ~ "batchA",
                           as.numeric(str_match(sample, "BCC(.*)")[,2]) %in% c(1:12)|grepl("HSJ", sample, ignore.case = F) ~ "batchB",
                           as.numeric(str_match(sample, "BCC(.*)")[,2]) %in% c(13:36) ~ "batchC"),
         sample_source = case_when(grepl("BCC", sample, ignore.case = T) ~ "BC_Children",
                                   grepl("hsj", sample, ignore.case = T) ~ "Montreal",
                                   TRUE ~ NA)) %>%
  arrange(batch)

# batch_key %>% group_by(batch) %>% summarize(n())

```









